Inattentive responding can induce spurious associations between task behavior and symptom measures

### 2021-04-09 manuscripts/main.tex
already ready for preprint

% make abstract
\abstract{
A common research design in the field of computational psychiatry involves leveraging the power of online participant recruitment to assess correlations between behaviour in cognitive tasks and the self-reported severity of psychiatric symptoms in large, diverse samples. Although large online samples have many advantages for psychiatric research, some potential pitfalls of this research design are not widely understood. Here we detail circumstances in which entirely spurious correlations may arise between task behaviour and symptom severity as a result of inadequate screening of careless or low-effort responding on psychiatric symptom surveys. Specifically, since many psychiatric symptom surveys have asymmetric ground-truth score distributions in the general population, participants who respond carelessly on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, and are not excluded from analysis, this may result in a spurious association between greater symptom scores and worse behavioural task performance. Here, we demonstrate exactly this pattern of results in $N = 386$ participants recruited online to complete a self-report symptom battery and a short reversal-learning choice task. We show that many behaviour-symptom correlations are entirely abolished when participants flagged for careless responding on surveys are excluded from analysis. We also show that exclusion based on task performance alone is not sufficient to prevent these spurious correlations. Of note, we demonstrate that false-positive rates for these spurious correlations \emph{increase} with sample size, contrary to common assumptions. We offer guidance on how researchers using this general experimental design can guard against this issue in future research; in particular, we recommend the adoption of screening methods for self-report measures that are currently uncommon in the field.
}

@@

With the advantages of online data collection also come specific drawbacks. Since participants recruited from online labour markets are typically completing experiments in their homes, they may be more likely to be distracted or multi-tasking during an experiment. Similarly, unsupervised participants may be more likely to use heuristic response strategies with the intention to minimise expenditure of time and cognitive effort (e.g., responding randomly on self-report surveys or behavioral tasks). Here, we will refer to such inattentive or low-effort behaviors as careless/insufficient effort (C/IE) responding \cite{huang2012detecting, curran2016methods}. Among researchers using online labour markets, a common view is that poor-quality data resulting from C/IE responding can simply be treated as a source of unsystematic measurement error that can be overcome with increased sample sizes \cite{gillan2016taking, chandler2020participant}. Similarly, it is common practice in online behavioral research to mitigate poor-quality data using the same screening methods that are typically used in in-person data collection (e.g., excluding participants who perform at- or below-chance on behavioral tasks).

Here we wish to draw attention to an underappreciated feature of online psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey accurately \cite{chandler2020participant, ophir2020turker}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening \cite{barends2019noncompliant, thomas2017validity, hauser2016attentive}. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. In a new dataset from two separate online labour marketplaces, we found that screening based on task behavior fails to adequately identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept dataset that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms of depression and anxiety that are abolished when data are screened properly.

@@

\begin{figure}[t!]
\includegraphics[width=16cm]{../figures/main_02a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Raincloud plots of total symptom scores in attentive (red) and C/IE (blue) participants. Each colored dot represents the symptom score for one participant. The black dots display the average score within each group (error bars denote 95\% bootstrap confidence interval). The scales are ordered according to their estimated skew from top-left (7-up) to bottom-right (PSWQ). The average level of symptom endorsement is most markedly different between groups in measures with lowest overall rates of endorsement.}
\label{fig:distributions}
\end{figure}

We hypothesise that spurious behavior-symptom correlations may emerge due to a mean-shift in the average level of symptom endorsement in participants engaging in C/IE responding relative to attentive participants. In turn, a mean-shift is expected to occur when the overall rate of symptom endorsement is low. To assess this, we visually inspected for each self-report measure the distribution of symptom scores in participants flagged by the infrequency measure vs. those that were not (Figure \ref{fig:distributions}). In line with our predictions, the average level of symptom endorsement was noticeably exaggerated in C/IE-responding participants for the symptom measures where symptom scores were most positively-skewed (7-up, 7-down, GAD-7). In contrast, where there was higher rates of symptom endorsement overall, the distributions of symptom scores between the two groups of participants were less noticeably distinct. Permutation testing confirmed that observed mean-shifts in symptom scores for C/IE participants were statistical significant for the majority of the symptom measures (Table \ref{tab:validation}).

\begin{table}[!t]
\centering
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lcccccccc}
\toprule
& & \multicolumn{3}{c}{Total Score} & \multicolumn{2}{c}{Cronbach's $\alpha$} & \multicolumn{2}{c}{\% Clinical Cutoff} \\
\cmidrule(lr){3-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
 Subscale & Skew & Attentive & C/IE & \textit{t}-value & Attentive & C/IE & Before & After \\
\midrule
7-up   &  0.81 &    3.87 &   10.15 &  -13.31^* &    0.84 &    0.84 &     13.0\% &       4.0\% \\
7-down &  0.76 &    4.75 &   10.68 &   -9.99^* &    0.94 &    0.88 &     17.4\% &       9.3\% \\
GAD-7  &  0.75 &    4.86 &    9.73 &   -7.88^* &    0.92 &    0.87 &     25.9\% &      17.3\% \\
BIS    &  0.78 &    7.74 &    7.92 &   -0.54 &    0.83 &    0.62 &        - &             - \\
BAS    &  0.17 &   15.73 &   16.16 &   -0.91 &    0.84 &    0.71 &        - &             - \\
SHAPS  &  0.26 &    8.02 &   10.85 &   -4.04^* &    0.90 &    0.81 &     17.9\% &      14.6\% \\
PSWQ   &  0.19 &    4.78 &    6.74 &   -4.78^* &    0.93 &    0.81 &      7.3\% &       7.0\% \\
\bottomrule
\end{tabular}
\captionsetup{width=0.88\textwidth}
\caption{Descriptive statistics of the self-report symptom measures between attentive and C/IE participants. Skew: the empirical skewness of the distribution of total symptom scores. Total score: the average symptom score between attentive and C/IE participants. * indicates statistical significance. Cronbach's $\alpha$: a measure of response consistency, where greater values closer to 1 indicate greater consistency in responses. \% Clinical Cutoff: the percentage of participants reaching threshold for clinical symptomology before and after screening based on the infrequency measure. (The BIS/BAS scales do not have clinical thresholds.)}
\label{tab:validation}
\end{table}

@@

\begin{figure}[!t]
\includegraphics[width=17cm]{../figures/main_03a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. Only statistically significant correlations are specified (not corrected for multiple comparisons). Black Xs indicate significant correlations ablated under screening. (The unthresholded correlations are included in the Supplement.) No Screening = no exclusions; Accuracy Only = exclusions based on chance-level performance in the reversal-learning task; Infrequency Only = exclusions based on invalid or improbable responses to infrequency items; Both Types = exclusions based on the previous two measures. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
\label{fig:correlations}
\end{figure}

@@

In stark contrast, when we rejected participants based on self-report screening  (removing 22\% of participants who endorsed one or more invalid or improbable responses on the infrequency items; Figure \ref{fig:correlations}C), the number of significant correlations was markedly reduced, particularly for several of the most skewed symptom measures (7-down, GAD-7) and proxy measures of task attentiveness (e.g. accuracy, inverse temperature). This pattern of correlations were largely similar when rejections were applied based on both task and self-report screening measures (Figure \ref{fig:correlations}D). We also note also that with stricter screening, the remaining significant correlations were, for the most part (but not always), weaker.

@@

%One possible objection to this interpretation is that the reduction in significant correlations following self-report screening and rejection was a result of the smaller sample size after removal of C/IE respondents (which comprised over 20\% of the sample). To test this alternative hypothesis, we compared the average (absolute) correlation size before and after infrequency-based exclusions were applied, restricted to correlations that were no longer significant afterward screening (the complete tables of correlations are included in the supplement). On average, we observed a significant reduction in correlation effect sizes ($\Delta \rho = -0.05, t = -2.87, p = 0.012$). following rejections, suggesting that the reduction in significant correlations following screening was unlikely to be driven solely by a reduction in statistical power.

One possible objection to this interpretation is that the reduction in significant correlations following- self-report screening and exclusion was a result of the reduced sample size after removal of C/IE respondents (which comprised over 20\% of the sample). To test this alternative hypothesis, we performed the same correlation analysis after removing random subsets of participants, fixing the sample size to that after excluding C/IE respondents. The rate of significant correlations in the bootstrapped samples are reported in Figure S\ref{fig:random}. Briefly, the pattern of significant correlations visually resembled the correlation structure prior to rejections (Figure \ref{fig:correlations}A), and statistical testing confirmed the pattern was more similar to that before screening than after screening using the infrequency measure (t = 262.49, p < 0.001). Thus, the reduction in significant correlations following screening was unlikely to be driven solely by a reduction in statistical power.

@@

\begin{figure}[!t]
\includegraphics[width=16cm]{../figures/main_03b.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{False positive rates for spurious correlations \emph{increase} with sample size. \textit{Left:} Bootstrapped Spearman correlations between learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. The grey error bars indicate 95\% bootstrap confidence intervals. The thick dashed lines indicate the threshold for statistical significance for the Spearman correlation at the corresponding sample size. Markers are jittered along the x-axis for legibility. \textit{Right:} False positive rates for learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. False positive rate was calculated as the proportion of bootstrap samples in which the correlation between $\kappa$ and 7-down was statistically significant.}
\label{fig:bootstrap}
\end{figure}

@@

\begin{figure}[!b]
\includegraphics[width=16cm]{../figures/main_03c.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Softmax regression decision weights in attentive (red) and C/IE (blue) participants. The weights dictate the extent to which the recent history of rewards, nonrewards, or previous choices influence current choice. Error bars indicate the 95\% highest density interval of each weight. * denotes where the 95\% highest density interval of the difference in weights excludes zero.}
\label{fig:regression}
\end{figure}

@@

This pattern of results also helps explain the pre-screening correlations with asymmetric learning rates. Previous work has established that, when choice hysteresis is not accounted for in reinforcement learning models, it can manifest as positive learning rate asymmetries \cite{Katahira2018-ke, sugawara2021dissociation}. Since C/IE participants showed increased hysteresis, which our reinforcement learning model did not explicitly account for, we should expect to find a negative correlation between learning-rate asymmetries and symptoms before C/IE participants are excluded. Indeed, this is what we observed above.

% This pattern of results, also mirrored in overall reduced outcome sensitivity, explains the pre-screening pattern of correlations with win-stay and lose-shift rates. But what of the learning rate asymmetries? This too can be explained by choice hysteresis. Previous work has established that, when choice hysteresis is not accounted for in reinforcement learning models, it can manifest as positive learning rate asymmetries \cite{Katahira2018-ke}. As such, given that attentive participants show increased repetition, which our reinforcement learning model did not explicitly account for, we should expect to find a negative (positive) correlation between positive (negative) learning rates and C/IE responding. Indeed, this is precisely the pattern of correlations observed above.

In sum, the theory-agnostic analysis of task behavior revealed that C/IE participants exhibited a qualitatively distinct behavioral strategy on the probabilistic reversal-learning task. C/IE participants showed less adaptive choice hysteresis. Moreover, they were numerically (but not significantly) less sensitive to outcome sensitivities. This latter finding helps clarify in part why we observed low correspondence between task and self-report screening measures (that is, C/IE participants were not significantly more likely to respond randomly during the task). These results also present another hidden danger of C/IE responding: qualitatively distinct patterns of behavior under C/IE responding can bias the estimation of parameters of theoretical interest if not properly accounted for.

%To conclude, the theory-agnostic analysis of task behavior revealed that careless participants exhibited less choice hysteresis and instead showed a pattern of responding more akin to win-stay lose-shift heuristic. This finding helps clarify in part why we observed low correspondence between task and self-report screening measures: on average, careless participants were not engaging in random responding (which would have been detected by the accuracy screening measure) but instead in satisficing behaviors. Most importantly, the results present another hidden danger of C/IE responding. When participants utilize heuristic decision-making, this may manifest in difficult-to-predict forms when computational models are misspecified. Thus, C/IE responding may not only drive spurious correlations but also spurious inferences about what behavior-symptom correlations entail.

@@

Viewed from the perspective of effort expenditure, our results suggest that participants appraised the cost/benefit trade-off differently for behavioral tasks and self-report surveys. Specifically, we found that only 7\% of participants were at chance-level performance in our (easy) task, compared to more than 22\% of participants who failed one or more (even easier) attention-check items in the self-report surveys. Moreover, different measures of C/IE responding were weakly or not at all correlated between task behavior and self-report responses. This suggests the motivation for effortful responding was greater in the behavioral task, though precisely why is unclear. One possibility is that we gave participants a monetary incentive for attentive responding only during the task (a common practice, according to our literature review). A second possibility is that participants expected fewer consequences for C/IE responding during the self-report surveys, a reasonable assumption in light of how infrequently previous experiments have screened self-report data. Alternatively, participants may have found the gamified behavioral task more engaging or the self-report inventory more tedious. Regardless of the reason, this discrepancy reinforces our observations concerning the inadequacy of behavioral-task screening as a stand-alone method for identifying C/IE responding. Since, in general, participants may appraise costs and benefits of effortful responding differently for behavioral tasks and self-report surveys, screening for C/IE responding on one data modality may in general be unsuitable for identifying it in the other. We therefore recommend screening on each component of an experiment.

One complicating factor for our argument is that C/IE responding may manifest in other ways than simply random responding for both behavioral tasks and self-report surveys. Indeed, in general there are more ways to respond carelessly than to respond attentively to a task or self-report inventory (e.g., random response selection, straight-lining, zig-zagging, acquiescence bias) \cite{curran2016methods}. Indeed, we observed qualitatively distinct behavioral strategy of decreased adaptive hysteresis in C/IE participants in the reversal learning task. The specific response strategy a participant adopts is likely to reflect the idiosyncratic integration of multiple perceived benefits (e.g. time saved, effort avoided) and costs (e.g. loss of performance bonuses, risk of detection and pay forfeited). As has been previously documented \cite{desimone2018dirty}, the presence of multiple response strategies makes it clear why certain screening measures are more or less likely to correlate. For example, the inter-item standard deviation and personal reliability measures are both sensitive to statistically random responding, but less sensitive to straight-lining. Most importantly, a diversity of heuristic response strategies highlights the need for many screening measures of C/IE responding, each sensitive to different heuristic strategies.

@@

One concern with performing rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias \cite{elwert2014endogenous}. That is, we have been treating a tendency towards C/IE responding as independent from psychopathology; however, to the extent that C/IE responding reflects lack of motivation \cite{barch2015mechanisms}, avoidance of effort \cite{cohen2001impairments, culbreth2016negative}, or more frequent lapses of attention \cite{kane2016individual, robison2017neurotic}, one might hypothesise a true underlying association between psychopathology and careless responding in online studies. In such cases, it is plausible that rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants.

Although, in theory, overcontrol bias may reduce power to detect legitimate symptom-behavior correlations, for overcontrol based on self-report questionnaires to seriously impact results it would have to be the case that symptomatic participants frequently endorsed improbable or impossible responses to infrequency-item checks (e.g., responding `Agree' to ``I competed in the 1917 Olympic Games"). We would argue that even if such participants truly are experiencing severe symptoms of motivation or attention, there is likely to be limited utility in measuring these symptoms using a self-report measure that they are unable to complete veridically. A similar rationale underlies the widespread use of semi-structured interviews and other clinician-report measures rather than self-report measures for in-clinic psychiatric research. We would argue that, if the psychiatric phenomenon being studied is such that this issue warrants concern, the research question may be better suited to an in-person study design involving participants in the clinic who meet full diagnostic criteria than a correlational design involving an online convenience sample.

Notwithstanding the above, one response to this legitimate concern is to take a graded approach to screening and excluding participants \cite{Kim2018-ev}. That is, participants could be screened with respect to a multitude of measures and only the consistently flagged participants be removed, thereby reducing the risk of inducing bias. Another possibility is to use sensitivity analysis as an alternative to exclusion, testing whether full-sample observed correlations are robust to the exclusion of participants flagged by measures of C/IE responding. We note that the strict screening approach used in the present study did not preclude us from identifying symptomatic participants or behavior-symptom correlations. Indeed, we found in our sample roughly 10\% of participants endorsing symptoms consistent with clinical levels of depression, and approximately 20\% consistent with clinical levels of acute anxiety. These estimates are within the realm of epidemiological norms \cite{kessler2012twelve, lowe2008validation, yarrington2021impact}.. We also observed some positive correlations between anxiety and choice behavior that were consistent with effects found in previous literature \cite{huang2017computational, harle2017anhedonia, garrett2018updating}. For example, we found higher lose-shift rates and higher learning rates following negative prediction errors correlated with self-reported anxiety. This suggests that the screening methods we employed were not so aggressive as to attenuate behavior-symptom correlations that would be expected from the literature.

>>>>

% make abstract
\abstract{
A common research design in the field of computational psychiatry involves leveraging the power of online participant recruitment to assess correlations between behaviour in cognitive tasks and the self-reported severity of psychiatric symptoms in large, diverse samples. Although large online samples have many advantages for psychiatric research, some potential pitfalls of this research design are not widely understood. Here we detail circumstances in which entirely spurious correlations may arise between task behaviour and symptom severity as a result of inadequate screening of careless or low-effort responding on psychiatric symptom surveys. Specifically, since many psychiatric symptom surveys have asymmetric ground-truth score distributions in the general population, participants who respond carelessly on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, and are not excluded from analysis, this may result in a spurious association between greater symptom scores and worse behavioural task performance. Here, we demonstrate exactly this pattern of results in $N = 386$ participants recruited online to complete a self-report symptom battery and a short reversal-learning choice task. We show that many behaviour-symptom correlations are entirely abolished when participants flagged for careless responding on surveys are excluded from analysis. We also show that exclusion based on task performance alone is not sufficient to prevent these spurious correlations. Of note, we demonstrate that false-positive rates for these spurious correlations \emph{increase} with sample size, contrary to common assumptions. We offer guidance on how researchers using this general experimental design can guard against this issue in future research; in particular, we recommend the adoption of screening methods for self-report measures that are currently uncommon in this field.
}

@@

With the advantages of online data collection also come specific drawbacks. Since participants recruited from online labour markets are typically completing experiments in their homes, they may be more likely to be distracted or multi-tasking during an experiment. They may also be more likely to use heuristic response strategies with the intention to minimise expenditure of time and cognitive effort (e.g., responding randomly on self-report surveys or behavioral tasks). Here, we will refer to such inattentive or low-effort behaviors as careless/insufficient effort (C/IE) responding \cite{huang2012detecting, curran2016methods}. Among researchers using online labour markets, a common view is that poor-quality data resulting from C/IE responding can simply be treated as a source of unsystematic measurement error that can be overcome with increased sample sizes \cite{gillan2016taking, chandler2020participant}. Common practice in online behavioral research is to mitigate poor-quality data using the same screening methods that are typically used in in-person data collection (e.g., excluding participants who perform at- or below-chance on behavioral tasks). However, these methods may be specifically inappropriate for online psychiatry studies, as we detail below.

Here we wish to draw special attention to an underappreciated feature of online psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey accurately \cite{chandler2020participant, ophir2020turker}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. In a new dataset from two separate online labour marketplaces, we found that screening based on task behavior fails to adequately identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept dataset that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms of depression and anxiety that are abolished when data are screened properly.

@@

\begin{figure}[t!]
\includegraphics[width=16cm]{../figures/main_02a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Raincloud plots of total symptom scores in attentive (red) and C/IE (blue) participants. Each colored dot represents the symptom score for one participant. Black circles: average score within each group (error bars denote 95\% bootstrap confidence interval). Shaded plots: estimated distribution of responses for each group of participants. The scales are ordered approximately according to their estimated skew (see Table 2) from top-left (7-up) to bottom-right (PSWQ). The average level of symptom endorsement is most markedly different between groups in measures with lowest overall rates of endorsement.}
\label{fig:distributions}
\end{figure}

We hypothesise that spurious behavior-symptom correlations may emerge due to a mean-shift in the average level of symptom endorsement in participants engaging in C/IE responding relative to attentive participants. In turn, a mean-shift is expected to occur when the overall rate of symptom endorsement is low. In line with our predictions, the average level of symptom endorsement was noticeably exaggerated in C/IE-responding participants for the symptom measures where symptom scores were most positively-skewed (7-up, 7-down, GAD-7; see Figure \ref{fig:distributions}). In contrast, where there was higher rates of symptom endorsement overall, the distributions of symptom scores between the two groups of participants were less noticeably distinct. Permutation testing confirmed that observed mean-shifts in symptom scores for C/IE participants were statistically significant for the majority of symptom measures (Table \ref{tab:validation}).

\begin{table}[!t]
\centering
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lcccccccc}
\toprule
& & \multicolumn{3}{c}{Total Score} & \multicolumn{2}{c}{Cronbach's $\alpha$} & \multicolumn{2}{c}{\% Clinical Cutoff} \\
\cmidrule(lr){3-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
 Subscale & Skew & Attentive & C/IE & \textit{t}-value & Attentive & C/IE & Before & After \\
\midrule
7-up   &  0.81 &    3.87 &   10.15 &  -13.31^* &    0.84 &    0.84 &     13.0\% &       4.0\% \\
7-down &  0.76 &    4.75 &   10.68 &   -9.99^* &    0.94 &    0.88 &     17.4\% &       9.3\% \\
GAD-7  &  0.75 &    4.86 &    9.73 &   -7.88^* &    0.92 &    0.87 &     25.9\% &      17.3\% \\
BIS    &  0.78 &    7.74 &    7.92 &   -0.54 &    0.83 &    0.62 &        - &             - \\
BAS    &  0.17 &   15.73 &   16.16 &   -0.91 &    0.84 &    0.71 &        - &             - \\
SHAPS  &  0.26 &    8.02 &   10.85 &   -4.04^* &    0.90 &    0.81 &     17.9\% &      14.6\% \\
PSWQ   &  0.19 &    4.78 &    6.74 &   -4.78^* &    0.93 &    0.81 &      7.3\% &       7.0\% \\
\bottomrule
\end{tabular}
\captionsetup{width=0.88\textwidth}
\caption{Descriptive statistics of the self-report symptom measures between attentive and C/IE participants. Skew: the empirical skewness of the distribution of total symptom scores. Total score: the average symptom score across attentive and C/IE participants. Stars indicate statistical significance at $p<0.05$. Cronbach's $\alpha$: a measure of response consistency, where values closer to 1 indicate greater consistency in responses. \% Clinical Cutoff: the percentage of participants reaching threshold for clinical symptomology before and after screening based on the infrequency measure. (The BIS/BAS scales do not have clinical thresholds.)}
\label{tab:validation}
\end{table}

@@

\begin{figure}[!t]
\includegraphics[width=17cm]{../figures/main_03a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. Only statistically significant correlations are shown ($p<0.05$ not corrected for multiple comparisons; all correlations are shown in the Supplement). Black Xs indicate significant correlations ablated under screening. No Screening = no exclusions; Accuracy Only = exclusions based on chance-level performance in the reversal-learning task; Infrequency Only = exclusions based on invalid or improbable responses to infrequency items; Both Types = exclusions based on the previous two measures. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
\label{fig:correlations}
\end{figure}

@@

In stark contrast, when we rejected participants based on self-report screening  (removing 22\% of participants who endorsed one or more invalid or improbable responses on the infrequency items; Figure \ref{fig:correlations}C), the number of significant correlations was markedly reduced, particularly for several of the most skewed symptom measures (7-down, GAD-7) and proxy measures of task attentiveness (e.g. accuracy, inverse temperature). This pattern of correlations was largely similar when rejections were applied based on both task and self-report screening measures (Figure \ref{fig:correlations}D). We also note that with stricter screening, the remaining significant correlations were, for the most part (but not always), weaker.

@@

One possible objection to this interpretation is that the reduction in significant correlations following self-report screening was a result of the reduced sample size after removal of C/IE respondents (which comprised over 20\% of the sample). To test this alternative hypothesis, we performed the same correlation analysis after removing random subsets of participants, fixing the sample size to that obtained after excluding C/IE respondents. In this case, the pattern of significant correlations was more similar to that before screening than after screening using the infrequency measure (t = 262.49, $p < 0.001$; Figure \ref{fig:random}, compare to Figure \ref{fig:correlations}A). Thus, the reduction in significant correlations following screening was unlikely to be driven solely by a reduction in statistical power.

@@

\begin{figure}[!t]
\includegraphics[width=16cm]{../figures/main_03b.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{False positive rates for spurious correlations \emph{increase} with sample size. \textit{Left:} Bootstrapped Spearman correlations between learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. The grey error bars indicate 95\% bootstrap confidence intervals. The thick dashed lines indicate the threshold for statistical significance for the Spearman correlation at the corresponding sample size. Markers are jittered along the x-axis for legibility. \textit{Right:} False positive rates for learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. False positive rate was calculated as the proportion of bootstrap samples in which the correlation between $\kappa$ and 7-down was statistically significant. The horizontal dotted line denotes the expected false positive rate at $\alpha = .05$.}
\label{fig:bootstrap}
\end{figure}

@@

\begin{figure}[!b]
\includegraphics[width=16cm]{../figures/main_03c.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Softmax regression decision weights in attentive (red) and C/IE (blue) participants. The weights dictate the extent to which the recent history of rewards, nonrewards, or previous choices influence current choice. Error bars indicate the 95\% highest density interval of each weight. * denotes where the 95\% highest density interval of the difference in weights excluded zero.}
\label{fig:regression}
\end{figure}

@@

This pattern of results also helps explain the pre-screening correlations with asymmetric learning rates. Previous work has established that, when choice hysteresis is not accounted for in reinforcement learning models, it can manifest as positive learning rate asymmetries \cite{Katahira2018-ke, sugawara2021dissociation}. Since C/IE participants showed decreased hysteresis, which our reinforcement learning model did not explicitly account for, we should expect to find a negative correlation between learning-rate asymmetries and symptoms before C/IE participants are excluded. Indeed, this is what we observed above.

In sum, the theory-agnostic analysis of task behavior revealed that C/IE participants exhibited a qualitatively distinct behavioral strategy on the probabilistic reversal-learning task. C/IE participants showed less adaptive choice hysteresis. Moreover, they were numerically (but not significantly) less sensitive to outcomes. The latter finding helps clarify in part why we observed low correspondence between task and self-report screening measures (that is, C/IE participants were not significantly more likely to respond randomly during the task). These results also present another hidden danger of C/IE responding: qualitatively distinct patterns of behavior under C/IE responding can bias the estimation of parameters of theoretical interest if not properly accounted for.

@@

Viewed from the perspective of effort expenditure, our results suggest that participants appraised the cost/benefit trade-off differently for behavioral tasks and self-report surveys. Specifically, we found that only 7\% of participants were at chance-level performance in our task, compared to more than 22\% of participants who failed one or more attention-check items in the self-report surveys. Moreover, different measures of C/IE responding were weakly or not at all correlated between task behavior and self-report responses. This suggests the motivation for effortful responding was greater in the behavioral task, though precisely why is unclear. One possibility is that we gave participants a monetary incentive for attentive responding only during the task (a common practice, according to our literature review). A second possibility is that participants expected fewer consequences for C/IE responding during the self-report surveys, a reasonable assumption in light of how infrequently previous experiments have screened self-report data. Alternatively, participants may have found the gamified behavioral task more engaging or the self-report inventory more tedious. Regardless of the reason, this discrepancy reinforces our observations concerning the inadequacy of behavioral-task screening as a stand-alone method for identifying C/IE responding. Since, in general, participants may appraise costs and benefits of effortful responding differently for behavioral tasks and self-report surveys, screening for C/IE responding on one data modality may in general be unsuitable for identifying it in the other. We therefore recommend screening on each component of an experiment.

One complicating factor for our argument is that C/IE responding may manifest in other ways than simply random responding for both behavioral tasks and self-report surveys. Indeed, there are more ways to respond carelessly than to respond attentively to a task or self-report inventory (e.g., random response selection, straight-lining, zig-zagging, acquiescence bias) \cite{curran2016methods}. Here we observed that C/IE participants failed to maintain previously correct choices after an occasional non-reward in the probabilistic reversal learning task, perhaps responding more locally based on the most immediate reward. The specific response strategy a participant adopts is likely to reflect the idiosyncratic integration of multiple perceived benefits (e.g. time saved, effort avoided) and costs (e.g. loss of performance bonuses, risk of detection and pay forfeited). As has been previously documented \cite{desimone2018dirty}, the presence of multiple response strategies makes it clear why certain screening measures are more or less likely to correlate. For example, the inter-item standard deviation and personal reliability measures are both sensitive to statistically random responding, but less sensitive to straight-lining. Most importantly, a diversity of heuristic response strategies highlights the need for many screening measures of C/IE responding, each sensitive to different heuristic strategies.

@@

One major concern with performing rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias \cite{elwert2014endogenous}. That is, we have been treating a tendency towards C/IE responding as independent from psychopathology; however, to the extent that C/IE responding reflects lack of motivation \cite{barch2015mechanisms}, avoidance of effort \cite{cohen2001impairments, culbreth2016negative}, or more frequent lapses of attention \cite{kane2016individual, robison2017neurotic}, one might hypothesise a true underlying association between psychopathology and careless responding in online studies. In such cases, it is plausible that rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants.

However, when using self-report questionnaires for screening, for overcontrol to seriously impact results it would have to be the case that symptomatic participants frequently endorse improbable or impossible responses to infrequency-item checks (e.g., responding `Agree' to ``I competed in the 1917 Olympic Games"). In this case, and even if such participants truly are experiencing severe symptoms of motivation or attention, there is likely to be limited utility in measuring these symptoms using a self-report measure that they are unable to complete veridically. A similar rationale underlies the widespread use of semi-structured interviews and other clinician-report measures rather than self-report measures for in-clinic psychiatric research. We would therefore argue that, if the psychiatric phenomenon being studied is such that this issue warrants concern, the research question may be better suited to an in-person study design involving participants in the clinic who meet full diagnostic criteria than a correlational design involving an online convenience sample.

Notwithstanding the above, one response to this legitimate concern is to take a graded approach to screening and excluding participants \cite{Kim2018-ev}. That is, participants could be screened with respect to a multitude of measures and only the consistently flagged participants be removed, thereby reducing the risk of inducing bias. Another possibility is to use sensitivity analysis as an alternative to exclusion, testing whether full-sample observed correlations are robust to the exclusion of participants flagged by measures of C/IE responding. We note that the strict screening approach used in the present study did not preclude us from identifying symptomatic participants or behavior-symptom correlations. Indeed, we found in our sample roughly 10\% of participants endorsing symptoms consistent with clinical levels of depression, and approximately 20\% consistent with clinical levels of acute anxiety. These estimates are within the realm of epidemiological norms \cite{kessler2012twelve, lowe2008validation, yarrington2021impact}. We also observed some positive correlations between anxiety and choice behavior that were consistent with effects found in previous literature \cite{huang2017computational, harle2017anhedonia, garrett2018updating}. For example, we found higher lose-shift rates and higher learning rates following negative prediction errors correlated with self-reported anxiety. This suggests that the screening methods we employed were not so aggressive as to attenuate behavior-symptom correlations that would be expected from the literature.
